A Constraint Solving Approach to Parikh Images of Regular Languages

### 2023-06-22 main.tex
Work on section 4. Still WIP!
Closes: #35
Closes: Ta bort 4.1 #13

\section{Projections on Parikh Images}\label{sec:generalised}

It is easy to see that the Parikh map~$\ParikhMap$ represents a
homomorphism from the (free) non-commutative monoid~$\Strings$ to the
(free) commutative monoid~$\Naturals^k$. As we are often not
interested in the full Parikh image, but only in some projections of
it. We consider \emph{arbitrary} homomorphisms
$\Map:\: \Sigma^* \to \Monoid$, where
$\Monoid = \left(X;\MonoidOp;0_{\Monoid}\right)$ is a commutative
monoid. We give several
examples of such projections on Parikh images later in this section.

Observe that every homomorphism~$\Map:\: \Sigma^* \to \Monoid$ can be
represented as the composition~$h' \circ \ParikhMap$, for some
homomorphism~$h' : \Naturals^k \to M$. One of the insights underlying our
approach is that it is more efficient to directly compute a
projection on the Parikh image~$\Map(\Language)$, than to first compute the
standard image~$\ParikhMap(\Language)$ followed by projection to
some property of interest.

%\subsection{Applications of generalised Parikh images}\label{sec:applications}

\subsection{Example~1: Reasoning about String Length}

One such simplifying homomorphism can express string
length, the problem that originally motivated our study of the Parikh
map. This mapping is relevant when solving constraints that combine
language membership as well as word length, for instance the formula
\begin{equation}\label{eq:stringLength}
x \in \Language_1 \wedge y \in \Language_2 \wedge |x| > |y|
\end{equation}
mentioned in the introduction. To solve this formula, let
$\Monoid = \Naturals$, and define the homomorphism~$L$ by $L(a) = 1$
for all characters~$a \in \Alphabet$. The length of a string
$s = s_1 \RepeatSum{\Concat} s_n$ is given by
$L(s) = \sum L(s_i) = 1 \RepeatSum{+} 1 = n$, and to solve
\eqref{eq:stringLength} we can instead solve the equi-satisfiable
formula~$\alpha \in L(\Language_1) \wedge \beta \in L(\Language_2)
\wedge \alpha > \beta$. This paper proposes efficient native
procedures to reason about membership constraints
like~$\alpha \in L(\Language_1)$, avoiding the computation of
the complete image~$L(\Language_1)$.

\subsection{Example~2: Generalised String Constraints with Integer
  Datatype}\label{sec:parikh-automata}

Parikh images are also applicable for deciding more general classes of
string constraints~\cite{ostrich-plus}. Consider, for instance, a
constraint involving the substring operation:
\begin{equation}
  \label{eq:substring}
  x \in \Language_1 \wedge \; 0 \leq n \leq m \leq |x| \wedge x[n:m] \in \Language_2
\end{equation}
in which $x$ is a string variable, $n, m$ are integer variables, and
$x[n:m]$ denotes the substring of $x$ starting at position~$n$ running
until position~$m$. This constraint belongs to an expressive fragment
of string logic that cannot be decided by most state-of-the-art string
solvers. To illustrate the decision procedure proposed in
\cite{ostrich-plus} for string constraints of this kind, suppose that
$\Language_2$ is defined by the regular
expression~$\mathtt{(ab|c)*}$. An automaton recognising this language
is shown in \cref{fig:parikh-automata} (left).

We can model \eqref{eq:substring} using the notion of a \emph{Parikh
  automaton}~\cite{parikh-automata,expressiveness}. A Parikh automaton
is an automaton in which transitions are labelled both with characters
from the alphabet~$\Alphabet$, and with offset vectors defining the
increments of a finite number of counters. This means that Parikh
automata recognise words over an extended
alphabet~$\Alphabet \times D$, where $D \subseteq \Naturals^d$ is a
finite set of increment vectors (notation as
in~\cite{expressiveness}). We use the symbols~$\pi_\Alphabet, \pi_D$
to denote projections to the first and the second component of a
composite letter~$(a, \Vector{v})$, respectively, and extend those
projections to words:
\begin{equation*}
  \pi_\Alphabet((a_1, \Vector{v}_1) \RepeatSum{\Concat} (a_k, \Vector{v}_k))
  ~=~ a_1 \RepeatSum{\Concat} a_k,
  \qquad
  \pi_D((a_1, \Vector{v}_1) \RepeatSum{\Concat} (a_k, \Vector{v}_k))
  ~=~ \Vector{v}_1 \RepeatSum{+} \Vector{v}_k.
\end{equation*}

\begin{definition}\label{def:parikh-automata} A Parikh automaton of dimension $d
  \geq 0$ is a pair $\Tuple{\Automaton, C}$, where
  $C \subseteq \Naturals^d$ is a semi-linear set (or, equivalently, a
  Presburger formula), and $\Automaton$ is a finite automaton with the
  alphabet $\Alphabet \times D$, where $D \subseteq \Naturals^d$. We
  say that $\Tuple{\Automaton, \varphi}$ recognises a
  word~$w \in \Alphabet^*$ if and only if the automaton has a run
  accepting an extended word~$w' \in (\Alphabet \times D)^*$ such that
  $\pi_\Alphabet(w') = w$ and $\pi_D(w') \in C$.
\end{definition}

\begin{figure}[t]
  \centering
  \raisebox{5ex}{
    \includegraphics[scale=0.9]{counter_automaton_pre}
    }
  \hfill
  \includegraphics[scale=0.9]{counter_automaton}
  \caption{An automaton recognising the language $\mathtt{(ab|c)*}$
    (left), and a Parikh automaton describing the pre-image of this
    language under the substring operation~$\cdot[n:m]$ (right). For
    sake of presentation, the Parikh automaton contains
    $\epsilon$-transitions, which could be eliminated in the standard
    way.}\label{fig:parikh-automata}
    \Description[An 2-state automaton (left) and a slightly larger automaton with counters (right).]{
      The rightmost automaton is the starts with an initial state with a self-loop on any character that increments both registers by one, followed by an epsilon transition to a state with a self-loop on c that increments the lower register by one. From that state, the automaton has an epsilon-transition into the accepting state, as well as a transition on a into a state with only a transition back again, both incrementing the lower register by one. The accepting state has a self-loop on any character that does nothing to the registers.
      }
    \end{figure}

    Applied to \eqref{eq:substring}, the decision procedure in
\cite{ostrich-plus} will construct a pre-image of $\Language_2$ under
the substring operation~$\cdot[n:m]$, and check whether this pre-image
is consistent with the constraint~$x \in \Language_1$. Because the
substring operation depends on the values of the integer
variables~$n, m$, a Parikh automaton of dimension~$2$ is a suitable
formalism to describe the pre-image, resulting
in the automaton in Fig.~\ref{fig:parikh-automata} (right). The first
component of the increment vectors is used to count the number of
letters eliminated in the beginning of the string (value~$n$), while
the second component records the beginning of the eliminated suffix
(value~$m$).

Denoting the language described by Fig.~\ref{fig:parikh-automata}
(right) as $\Language_{pre}$, we can then replace \eqref{eq:substring}
with an equi-satisfiable formula that no longer contains any explicit
substring operation:
%
\begin{equation}
  \label{eq:parikh-constraint}
  p \in \Language_{pre} \wedge~
  \pi_\Alphabet(p) \in \Language_1
  \wedge~~ \pi_D(p) =
  \begin{bmatrix}
    n \\ m
  \end{bmatrix}~~
  \wedge 0 \leq n \leq m \leq |p|
\end{equation}
%
To check the satisfiability of \eqref{eq:parikh-constraint}, we need a
decision procedure that can process intersections of regular languages
(in this case, of $\Language_{pre}$ and $\Language_1$, synchronising
on $\Alphabet$), while imposing the side
condition~$0 \leq n \leq m \leq |p|$ on the increment sum. In
\cite{ostrich-plus}, this decision procedure turned out to be main
bottleneck of the string solver, which was one of the motivations to
develop the lazy algorithm proposed in this paper.

\iffalse
\begin{figure}[t]
  \centering
      \includegraphics[width=0.5\linewidth]{parikh_automaton}
      \caption{The automaton part of a Parikh automaton for $\AcaOrBc{}$ with
      $\Alphabet = \Set{\text{a, b, c}}, d = 3$. The semilinear set/Presburger
      formula containing the constraints on the final register values cannot be
      visualised.}\label{fig:parikh-automaton}
    \end{figure}
\fi

>>>>

\section{Projections on Parikh Images}\label{sec:generalised}
\input{generalised.tex}
